Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-10-04 paper/sections/preliminaries.tex
[ paper ] minor changes.
[ paper ] preliminaries reviewed.

\textit{Propositions.}
We define the classical propositional logic as usual.
A proposition is an expression of indivisible propositional atoms
(e.g., symbols $φ₀, φ₁, \cdots$),
binary logic connectives ($∧, ∨, ⇒, ⇔)$,
propositional negation $(¬)$, and logic constants $⊥$
and $⊤$. We use the inductive definition of the set \Prop
presented in \cite{VanDalen1994} as the mechanism to produce any
propositional formula.
Then, $ϕ_i \in \Prop$ for all $i \in \mathbb{N}$.
If $φ, ψ\in \Prop$, then $(φ\ \square\ ψ) \in \Prop$,
where $(φ\ \square\ ψ)$ stands for a proposition that use a
binary connective in the place for the $\square$ symbol.
If $φ \in \Prop$, then $(¬ φ) \in \Prop$.
Finally, $⊥, ⊤ \in \Prop$.

\emph{Derivations.}
A derivation of $φ$ is a mechanism to obtain the conclusion $φ$ from
the application of some deduction steps. Such a deduction step has
the form of a set of propositional \emph{premises} that conclude the
result: one or more propositions. As we show in the following
figure, the premises are always above the line, the conclusions
are below it, and the name of the inference or the theorem used is
on the right the line. As a notation convention, if the name of a
rule used in the derivation step is typed using verbatim, that
means, it is available in our implementation (see for instance,
theorems in \cite{AgdaProp}) used in some derivations latter.

\[
\label{eq:derivation}
\begin{bprooftree}
\AxiomC{φ}
\AxiomC{ψ}
\BinaryInfC{φ ∧ ψ}
\end{bprooftree}\qquad
\begin{bprooftree}
\AxiomC{[¬ φ]}
\noLine
\UnaryInfC{$\vdots$}
\noLine
\UnaryInfC{⊥}
\RightLabel{\abbre{RAA}}
\UnaryInfC{φ}
\end{bprooftree}
\]

@@

\label{eq:PEM-RAA}
\[
\begin{bprooftree}
\AxiomC{\ }
\RightLabel{\scriptsize\tt PEM}
\UnaryInfC{Γ ⊢ φ ∨ ¬ φ}
\end{bprooftree}
\qquad
\begin{bprooftree}
\AxiomC{Γ, ¬ φ ⊢ ⊥}
\RightLabel{\scriptsize\tt RAA}
\UnaryInfC{Γ ⊢ φ}
\end{bprooftree}
\qquad
\]

\emph{Natural Deductions.}
The natural deduction designed by Gentzen \improvement{Cite the work
by Gentzen.} defines a formal system for derivations that uses a set
of inference rules as the only way to obtain a conclusion.
We use this system to refer to formal proofs as natural deduction
proofs and conversely, in the latter.
We define our inference rules for \CPL in the Fig.
\ref{fig:CPL-inference-rules}. These rules give us the complete
meaning of the connectives; how to introduce them and
how they can be remove them in the conclusion side of the theorems.
\change{the size of the text.}

@@

\emph{Proof Trees.} We use trees to represent derivations and
natural deduction proofs (see for instance, the proof tree in a
\Metis' derivation in Fig.~\ref{fig:metis-example}).
In a proof tree, the root is the conclusion of the entire
derivation. The nodes labeled with the name of the inference rule
contains the result of applying that rule to their parents. Lastly,
the leaves in the tree become the premises, not all necessary
canceled (see a better discussion on cancellation of hypotheses in
\cite{VanDalen1994}).

>>>>

\textit{Propositions.}
We define the classical propositional logic as usual.
A proposition is an expression of indivisible propositional atoms
(e.g., symbols $φ₀, φ₁, \cdots$),
binary logic connectives ($∧, ∨, ⇒, ⇔)$,
propositional negation $(¬)$, and logic constants $⊥$
and $⊤$. We use the inductive definition of the set \Prop
presented in \cite{VanDalen1994} as the mechanism to produce any
propositional formula.
%Then, $ϕ_i \in \Prop$ for all $i \in \mathbb{N}$.
$⊥, ⊤ \in \Prop$.
If $φ, ψ\in \Prop$, then $(φ\ \square\ ψ) \in \Prop$,
where $(φ\ \square\ ψ)$ stands for a proposition that use the $\square$ symbol as one of the binary connectives mentioned above.
If $φ \in \Prop$, then $(¬ φ) \in \Prop$.

\emph{Derivations.}
A derivation of $φ$ is a mechanism to obtain the conclusion $φ$ from
the application of some deduction steps. Such a deduction step has
the form of a set of propositional \emph{premises} that conclude
one or more proposition as the conclusion.

As we show in the following
figure, the premises are always above the line, the conclusions
are below it, and the name of the inference or the theorem used is
on the left or on the right of the line. As a notation convention, if the name of a rule used in the derivation step is typed using verbatim font face, it means, the inference rule belongs to our implementation~\cite{AgdaProp, AgdaMetis} in the Proof Assistant of a inference rule presented
with a similar name.

\begin{equation*}
\label{eq:derivation}
\begin{bprooftree}
  \AxiomC{$φ$}
  \AxiomC{$ψ$}
  \RightLabel{∧-intro}
  \BinaryInfC{$φ ∧ ψ$.}
\end{bprooftree}
% \qquad
% \begin{bprooftree}
%   \AxiomC{$[¬ φ]$}
%   \noLine
%   \UnaryInfC{$\vdots$}
%   \noLine
%   \UnaryInfC{$⊥$}
%   \RightLabel{\abbre{RAA}}
%   \UnaryInfC{φ}
% \end{bprooftree}
\end{equation*}

@@

\begin{equation*}
\label{eq:PEM-RAA}
\begin{bprooftree}
\AxiomC{}
\RightLabel{PEM}
\UnaryInfC{$Γ ⊢ φ ∨ ¬ φ$,}
\end{bprooftree}\qquad
\begin{bprooftree}
\AxiomC{$Γ, ¬ φ ⊢ ⊥$}
\RightLabel{RAA}
\UnaryInfC{$Γ ⊢ φ$.}
\end{bprooftree}
\end{equation*}

\emph{Natural Deductions.}
The natural deduction designed by Gerhard Gentzen in 1930s defines a formal system for derivations that uses a set
of inference rules as the only way to obtain a conclusion.
We use this system to refer to formal proofs as natural deduction
proofs and conversely.
We define the \CPL formal system in the
Fig.~\ref{fig:CPL-inference-rules}. The rules described in the aforementioned figure give us the complete
meaning of the connectives; how to introduce them and
how they can be remove them in the conclusion side of the theorems.

@@

\emph{Proof Trees.} We use trees to represent derivations and
natural deduction proofs (see for instance, the proof tree in a
\Metis' derivation in Fig.~\ref{fig:metis-example}).
In a proof tree, the root is the conclusion of the entire
derivation. The nodes labeled with the name of the inference rule
contains the result of applying that rule to their parents. Lastly,
the leaves in the tree become the premises, not all necessary
canceled (see a better discussion about canceling  hypotheses in proof trees in Section 2.4 in \citeauthor{VanDalen1994}~\cite{VanDalen1994}).

### 2017-10-04 paper/sections/proof-reconstruction.tex
[ paper ] preliminaries reviewed.

Thus, we define the function
$unshunt_{1} : \Prop \to \Nat \to \Prop$
by structural recursion in its second argument in
Eq.~\ref{eq:unshunt-structured}.

@@

As we mentioned above, the natural number in the second argument of the
function used in Eq.~\ref{eq:unshunt-structured} stands for a complexity
measure of the function, in this case $unshunt_{1}$ function.

@@

Recall, we want to maintain the same outputs as the original unshunt
does in Eq.~\ref{eq:unshunt-definition} and the complexity measure of it,
is a bound for the number of times this function can call itself.

Therefore, we define an appropriate complexity measure
$unshunt_{cm}: \Prop → \NN$ for our $unshunt_{1}$ function presented in
Eq.~\ref{eq:unshunt-rank-definition}.

@@

Finally, the curated version of $unshunt_{0}$ function is presented
in Eq.~\ref{eq:unshunt-fixed}.

@@

In the following, we show how Theorem.~\ref{thm:thm-strip} proves the correctness of the process mentioned above. But first, let prove some lemmas for $unshunt$ function presented in Eq.~\ref{eq:unshunt-fixed},
and $split$ function from the definition in Eq.~\ref{eq:split-definition}.

@@

\begin{proof} Use induction on the cases defined by the outcome of the
$unshunt$ function in Eq.~\ref{eq:unshunt-structured}.
When $n = 0$, by definition, $Γ ⊢ φ$.
When $n = 1$, we apply the following theorem from~\cite{AgdaProp}

@@

Now, let us consider the $split$ function presented in
Eq.~\ref{eq:split-definition} that yields a conjunction of subgoals from a proposition, the goal in \Metis.

@@

\begin{proof} We now proceed by induction on the structure of the
formula φ in the definition of Eq.~\ref{eq:split-definition}.
For a complete proof, we refer the reader to~\cite{AgdaMetis}.\\[3mm]

@@

\begin{proof}
When φ ≡ ψ, from $Γ ⊢ conjunct(φ, ψ)$ by normalization, we get $Γ ⊢ ψ$.
By using $subst$ theorem Eq.~\ref{eq:substitution-theorem}, we get the desire conclusion. If the proposition φ is a conjunction, and $ψ ≡ conjunct(φ_{i}, ψ)$
for some $i = 1,\ 2$, then,

@@

First, we define the $rdisj$ function in Eq.~\ref{eq:rdisj-def} to convert a disjunction to be a right-associative disjunction, and following
a similar approach we define the $rconj$ function for conjunctions.

@@

The following function, $build_{∨} : \Prop → \Prop → \Prop$ in Eq.~\ref{eq:build-or}, intends to construct a disjunction from another disjunction, specifically, we want to rearrange the disjuncts in the source formula to match with the target disjunction for syntactic
equivalence.

@@

The function $factor : \Prop → \Prop$  in Eq.~\ref{eq:factor-definition} is intended to work only for right-associative disjunctions, and reduces a disjunction of the same disjunct (e.g., $factor(φ ∨ (φ ∨ \cdots)) \mapsto\ φ$).

@@

Now, with a right-associative disjunction by applying $rdisj$ function to
a proposition $φ$, $sbuild_{∨}: \Prop → \Prop → \Prop$ in
Eq.~\ref{eq:strong-build-or-definition} aims to construct the disjunction
ψ by using each of its disjuncts from the proposition φ. This approach to
build a disjunction from another is inspired on the description in
\citeauthor{bohme2010}~\cite{bohme2010} to prove equality between nested
disjunctions. \change{Make clear this, It's not clear the difference with build-or and sbuild-or.}

@@

\begin{theorem}[thm-clausify]
\label{thm:thm_clausify}
  $Γ ⊢ φ$ and $ψ \in\Prop$ then $Γ ⊢ clausify(φ, ψ)$, where,
  \begin{align}
      clausify(φ,ψ) =
      \begin{cases}
        ψ, &\text{ if }φ≡ψ\\
        reorder_{∧∨}(cnf(φ), ψ), &\text{otherwise.}
      \end{cases}
  \end{align}
\end{theorem}

\begin{proof} When $φ ≡ ψ$, we use the $subs$ theorem Eq.~\ref{eq:substitution-theorem}. Otherwise, we use Theorem \ref{thm:thm-reorder-and-or} and Theorem \ref{thm:thm-cnf}.
\end{proof}

@@

Finally, we remove redundancies in a proposition assuming that it is given in conjunctive normal form and right-associative, by composing the functions described above in the Eq.~\ref{eq:redun-ast}
for the function $redun^{*}$.

>>>>

Thus, we define the function
$unshunt_{1} : \Prop \to \Nat \to \Prop$
by structural recursion in its second argument in
Eq.~(\ref{eq:unshunt-structured}).

@@

As we mentioned above, the natural number in the second argument of the
function used in Eq.~(\ref{eq:unshunt-structured}) stands for a complexity
measure of the function, in this case $unshunt_{1}$ function.

@@

Recall, we want to maintain the same outputs as the original unshunt
does in Eq.~(\ref{eq:unshunt-definition}) and the complexity measure of it,
is a bound for the number of times this function can call itself.

Therefore, we define an appropriate complexity measure
$unshunt_{cm}: \Prop → \NN$ for our $unshunt_{1}$ function presented in
Eq.~(\ref{eq:unshunt-rank-definition}).

@@

Finally, the curated version of $unshunt_{0}$ function is presented
in Eq.~(\ref{eq:unshunt-fixed}).

@@

In the following, we show how Theorem.~\ref{thm:thm-strip} proves the correctness of the process mentioned above. But first, let prove some lemmas for $unshunt$ function presented in Eq.~(\ref{eq:unshunt-fixed}),
and $split$ function from the definition in Eq.~(\ref{eq:split-definition}).

@@

\begin{proof} Use induction on the cases defined by the outcome of the
$unshunt$ function in Eq.~(\ref{eq:unshunt-structured}).
When $n = 0$, by definition, $Γ ⊢ φ$.
When $n = 1$, we apply the following theorem from~\cite{AgdaProp}

@@

Now, let us consider the $split$ function presented in
Eq.~(\ref{eq:split-definition}) that yields a conjunction of subgoals from a proposition, the goal in \Metis.

@@

\begin{proof} We now proceed by induction on the structure of the
formula φ in the definition of Eq.~(\ref{eq:split-definition}).
For a complete proof, we refer the reader to~\cite{AgdaMetis}.\\[3mm]

@@

\begin{proof}
When φ ≡ ψ, from $Γ ⊢ conjunct(φ, ψ)$ by normalization, we get $Γ ⊢ ψ$.
By using $subst$ theorem Eq.~(\ref{eq:substitution-theorem}), we get the desire conclusion. If the proposition φ is a conjunction, and $ψ ≡ conjunct(φ_{i}, ψ)$
for some $i = 1,\ 2$, then,

@@

First, we define the $rdisj$ function in Eq.~(\ref{eq:rdisj-def}) to convert a disjunction to be a right-associative disjunction, and following
a similar approach we define the $rconj$ function for conjunctions.

@@

The following function, $build_{∨} : \Prop → \Prop → \Prop$ in Eq.~(\ref{eq:build-or}), intends to construct a disjunction from another disjunction, specifically, we want to rearrange the disjuncts in the source formula to match with the target disjunction for syntactic
equivalence.

@@

The function $factor : \Prop → \Prop$  in Eq.~(\ref{eq:factor-definition}) is intended to work only for right-associative disjunctions, and reduces a disjunction of the same disjunct (e.g., $factor(φ ∨ (φ ∨ \cdots)) \mapsto\ φ$).

@@

Now, with a right-associative disjunction by applying $rdisj$ function to
a proposition $φ$, $sbuild_{∨}: \Prop → \Prop → \Prop$ in
Eq.~(\ref{eq:strong-build-or-definition}) aims to construct the disjunction
ψ by using each of its disjuncts from the proposition φ. This approach to
build a disjunction from another is inspired on the description in
\citeauthor{bohme2010}~\cite{bohme2010} to prove equality between nested
disjunctions. \change{Make clear this, It's not clear the difference with build-or and sbuild-or.}

@@

\begin{theorem}[thm-clausify]
\label{thm:thm-clausify}
  $Γ ⊢ φ$ and $ψ \in\Prop$ then $Γ ⊢ clausify(φ, ψ)$, where,
  \begin{align}
      clausify(φ,ψ) =
      \begin{cases}
        ψ, &\text{ if }φ≡ψ\\
        reorder_{∧∨}(cnf(φ), ψ), &\text{otherwise.}
      \end{cases}
  \end{align}
\end{theorem}

\begin{proof} When $φ ≡ ψ$, we use the $subs$ theorem Eq.~(\ref{eq:substitution-theorem}). Otherwise, we use Theorem \ref{thm:thm-reorder-and-or} and Theorem \ref{thm:thm-cnf}.
\end{proof}

@@

Finally, we remove redundancies in a proposition assuming that it is given in conjunctive normal form and right-associative, by composing the functions described above in the Eq.~(\ref{eq:redun-ast})
for the function $redun^{*}$.
